Parallel Hierarchical Clustering using \\ Rank-Two Nonnegative Matrix Factorization \thanks{Identify applicable funding agency here. If none, delete this.}

### 2020-07-16 latex/paper.tex
sections 2 and 3 comments

Nonnegative Matrix Factorization (NMF) has been demonstrated to be an effective tool for unsupervised learning problems including clustering \cite{}. \GB{add citations}
An NMF consists of two tall-and-skinny non-negative matrices whose product approximates a nonnegative data matrix.
That is, given an $m\times n$ data matrix $\M{A}$, we seek nonnegative matrices $\M{W}$ and $\M{H}$ that each have $k$ columns so that $\M{A} \approx \M{W}\M{H}^\Tra$.
Each pair of corresponding columns of $\M{W}$ and $\M{H}$ form a latent component of the NMF.
If the rows of $\M{A}$ correspond to features and the columns to samples, the $i$th row of the $\M{H}$ matrix represents the loading of sample $i$ onto each latent component and provides a soft clustering.
Because the $\M{W}$ factor is also nonnegative, each column can typically be interpreted as a latent feature vector for each cluster.

While standard NMF provides a flat clustering, NMF can also be used as a tool for hierarchical clustering.
Kuang and Park \cite{KP13} propose a method that uses rank-2 NMF to recursively bipartition the samples.
An NMF with $k=2$ determines a row of $\M{H}$ of length 2 for each sample that determines its part membership.
By recursively bipartitioning the samples in each part, the method determines a binary tree such that each sample belongs to a unique leaf and internal nodes of the tree determine higher-level clusters of the samples.
A single latent feature vector for each node can also be used for interpreting clusters.
Du et al. \cite{DKDP17} show that the latent features vectors of the leaves can also be used to initialize a $\M{W}$ within an algorithm for flat NMF, yielding fast convergence and overall speedup compared to initializing the algorithm with random data.

We illustrate the output of the hierarchical clustering method with an example data set and output tree.
Following Gillis et al. \cite{GKP15}, we apply the method to a hyperspectral imaging (HSI) data set of the Washington, D.C~national mall, which has pixel dimensions $1280 \times 307$ and 191 spectral bands.
\Cref{fig:dc} visualizes the output tree with 6 leaves along with their hierarchical relationships.
The root node, labeled 0, is a flattening of the HSI data to a 2D grayscale image.
Each other node is represented by an overlay of the member pixels of the clusters (in blue) on the original grayscale image.
The first bipartitioning separates vegetation (cluster 1) from non-vegetation (cluster 2), the bipartitioning of cluster 1 separates grass (cluster 3) from trees (cluster 4), the bipartitioning of cluster 2 separates buildings (cluster 5) from sidewalks/water (cluster 6), and so on.

While the hierarchical clustering method offers advantages in terms of interpretation as well as execution time compared to flat NMF, implementations of the algorithm are limited to single workstations and the dataset must fit in the available memory.
Currently available implementations can utilize multiple cores via MATLAB \cite{KP13} or explicit shared-memory parallelization \cite{SmallK}.
The goal of this work is to use distributed-memory parallelism to scale the algorithm to large datasets that require the memory of multiple compute nodes and to high processor counts.

\begin{itemize}
	\item discuss limitation of current implementations of the hierarchical method in MATLAB and shared-memory parallelization in terms of scalability to data size and bandwidth limitations
	\item brief summary of parallelization approach, point to analysis
	\item summarize results, highlight 600 GB data set and scaling to 2000-2500 cores, speedups of up to X for rank-2 NMF and up to Y for full hierarchical clustering algorithm
\end{itemize}

\GB{we should have perhaps compared against SmallK and also against flat NMF...}

\begin{figure}
\input{fig/dc}
\caption{Hierarchical Clustering of DC Mall Hyperspectral Image}
\label{fig:dc}
\end{figure}

@@

When the rank of the factorization (the number of columns of $\M{W}$ and $\M{H}$) is 2, the NNLS subproblems can be solved much more quickly because the number of possible active sets is only 4.
As explained in more detail in \Cref{sec:r2nmf}, the optimal solution across the 4 possibilities can be determined efficiently to solve the NNLS subproblem more quickly than general-rank approaches like block principal pivoting.
Because of the relative ease of solving the NMF problem for the rank-2 case, Kuang and Park \cite{KP13} propose a recursive method to use a rank-2 NMF to partition the input data into 2 parts, whereby each part can be further partitioned via rank-2 NMF of the corresponding original data.
This approach yields a hierarchical factorization, potentially uncovering more global structure of the input data and allowing for better scalability of the algorithm to large NMF ranks.

@@

All of the interprocessor communication in the algorithms presented in this work are encapsulated in collective communication operations that involve the full set of processors.
Algorithms for implementing the collective operations are built out of pairwise send and receive operations, and we assume the most efficient algorithms are used in our analysis \cite{TRG05,CH+07}.
The collectives used in our algorithms are all-reduce, all-gather, and reduce-scatter.
In an all-reduce, all processors start out with the same amount of data and all end with a copy of the same result, which is in our case a sum of all the inputs (and the same size as a single input).
The cost of an all-reduce of size $n$ words is $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n)$ for $n>p$ and $(\alpha+\beta) \cdot O(\log p)$ for $n<p$.
In an all-gather, all processors start out with separate data and all end with a copy of the same result, which is the union of all the input data.
If each processor starts with $n/p$ data and ends with $n$ data, the cost of the all-gather is $\alpha \cdot O(\log p) + \beta \cdot O(n)$.
In a reduce-scatter, all processors start out with the same amount of data and all end with a subset of the result, which is in our case a sum of all the inputs (and is smaller than its input).
If each processor starts with $n$ data and ends with $n/p$ data, the cost of the reduce-scatter is $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n)$.
In the case of all-reduce and reduce-scatter, the computational cost is typically dominated by the bandwidth cost because $\beta \gg \gamma$.

@@

As shown by Kuang and Park \cite{KP13}, determining which of the four possible solutions is feasible and optimal can be done efficiently by exploiting the following properties:
\begin{itemize}
	\item if the solution to the unconstrained least squares problem admits two positive values, it is the optimal solution to the nonnegatively constrained problem,
	\item if $\M{W}$ and $\M{A}$ are both nonnegative, then the candidate solution with two zero values is never (uniquely) optimal and can be discarded, and
	\item if the solution to the unconstrained problem does not admit two positive values, the better of the two remaining solutions can be determined by comparing $\V{a}_{j}^\Tra\V{w}_1 / \|\V{w}_1\|$ and $\V{a}_{j}^\Tra\V{w}_2 / \|\V{w}_2\|$.
\end{itemize}
If the unconstrained problem is solved via the normal equations, then the temporary matrices computed for the normal equations ($\M{W}^\Tra\M{W}$ and $\M{A}^\Tra\M{W}$) can be re-used to determine the better of the two solutions with a single positive variable.

@@

Given that the computational complexity of \Cref{alg:r2nnls} is $O(n)$ (or $O(m)$ when computing $\M{W}$), and the complexity of computing $\M{W}^\Tra\M{W}$ and $\M{H}^\Tra\M{H}$ is $O(m+n)$, the typical dominant cost of each iteration of Rank-2 NMF is that of computing $\M{A}^\Tra\M{W}$ and $\M{A}\M{H}$, which is $O(mn)$ when $\M{A}$ is dense and $O(\text{nnz}(A))$ when $\M{A}$ is sparse.

@@

The splitting procedure is specified in \Cref{alg:split}.
After the Rank-2 NMF is performed, the $\M{H}$ factor is used to determine part membership, and the columns of the $\M{W}$ factor are assigned to the child nodes.
The score of the node is computed as the reduction in overall NMF error if the node is split, which can be computed from the principal singular values of the subsets of columns of the node and its children.
The principal singular values of the children are computed via the power method (the principal singular value of the node was computed when its parent's score was computed).

@@

Given the 1D distribution, only one communication collective is required for the pair of matrix-vector multiplications.
That is, the approximate right singular vector $\V{v}$ is redundantly owned on each processor, and the approximate left singular vector $\V{u}$ is distributed across processors.
Each processor can compute its local $\V[\hat]{u}$ from $\V{v}$ without communication and use the result for its contribution to $\V{v}=\M{A}^\Tra \V{u}$.
An all-reduce collective is used to obtain a copy of $\V{v}$ on every processor for the next iteration, and the norm is redundantly computed without further communication.
Again, we use a fixed number of iterations for benchmarking, but the more practical stopping criterion is based on the relative change in $\sigma$.

@@

To analyze the overall cost of the hierarchical clustering algorithm, we sum the costs over all nodes in the tree.
Because the shape of the tree is data dependent and affects the overall costs, for the sake of analysis we will analyze only complete levels.
The number of rows in any node is $m$, the same as the root node, as each splitting corresponds to a partition of the columns.
Furthermore, because each split is a partition, every column of $\M{A}$ is represented exactly once in every complete level of the tree.
If we assume that all nodes perform the same number of NMF iterations ($\imath$) and power method iterations $(\jmath)$, then the dominating costs of a node with $\overline{n}$ columns is
\begin{equation*}
\label{eq:nodecost}
\begin{split}
\gamma \cdot O\left( \frac{(\imath+\jmath) m\overline{n} + \imath (m+\overline{n}) }{p} \right) + \beta \cdot O((\imath+\jmath) \overline{n}) \\ + \alpha \cdot O((\imath+\jmath) \log p).
\end{split}
\end{equation*}
Because the sum of the number of columns across any level of the tree is $n$, the cost of the $\ell$th level of the tree is
\begin{equation}
\label{eq:levelcost}
\begin{split}
\gamma \cdot O\left( \frac{(\imath+\jmath) mn + \imath m 2^\ell}{p} \right) + \beta \cdot O((\imath+\jmath) n) \\ + \alpha \cdot O((\imath+\jmath) 2^\ell \log p).
\end{split}
\end{equation}
Note that the only cost that depends on the level index $\ell$ is the latency cost.
Summing over levels and assuming the tree is nearly balanced and has height $O(\log k)$ where $k$ is the number of frontier nodes, we obtain an overall cost of \Cref{alg:hiernmf2} of
\begin{equation}
\label{eq:treecost}
\begin{split}
\gamma \cdot O\left( \frac{(\imath+\jmath) mn}{p}  \log k + \frac{\imath mk}{p} \right) + \beta \cdot O((\imath+\jmath) n \log k) \\ + \alpha \cdot O((\imath+\jmath) k \log p).
\end{split}
\end{equation}

>>>>

\GB{Prof Park, can you draft an introduction?  We are focusing on the dense case and using HSI and image clustering as applications.  Section 4 is still unfinished, but nearly all of the performance plots are there.  Feel free to leave placeholders for details that aren't obvious, we can clean them up Wednesday night.}

@@

When the rank of the factorization (the number of columns of $\M{W}$ and $\M{H}$) is 2, the NNLS subproblems can be solved much more quickly because the number of possible active sets is only 4.
As explained in more detail in \Cref{sec:r2nmf}, the optimal solution across the 4 possibilities \LM{sets} can be determined efficiently to solve the NNLS subproblem more quickly than general-rank approaches like block principal pivoting.
Because of the relative ease of solving the NMF problem for the rank-2 case, Kuang and Park \cite{KP13} propose a recursive method to use a rank-2 NMF to partition the input data into 2 parts, whereby each part can be further partitioned via rank-2 NMF of the corresponding original data.
This approach yields a hierarchical factorization, potentially uncovering more global structure of the input data and allowing for better scalability of the algorithm to large NMF ranks.

@@

All of the interprocessor communication in the algorithms presented in this work are encapsulated in collective communication operations that involve the full set of processors.
Algorithms for implementing the collective operations are built out of pairwise send and receive operations, and we assume the most efficient algorithms are used in our analysis \cite{TRG05,CH+07}.
The collectives used in our algorithms are all-reduce, all-gather, and reduce-scatter.
In an all-reduce, all processors start out with the same amount of data and all end with a copy of the same result, which is in our case a sum of all the inputs (and the same size as a single input).
The cost of an all-reduce of size $n$ words is $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n)$ for $n>p$ and $(\alpha+\beta) \cdot O(\log p)$ for $n<p$. \LM{is the n<p case right? where's the computation?}
In an all-gather, all processors start out with separate data and all end with a copy of the same result, which is the union of all the input data.
If each processor starts with $n/p$ data and ends with $n$ data, the cost of the all-gather is $\alpha \cdot O(\log p) + \beta \cdot O(n)$.
In a reduce-scatter, all processors start out with the same amount of data and all end with a subset of the result, which is in our case a sum of all the inputs (and is smaller than its input).
If each processor starts with $n$ data and ends with $n/p$ data, the cost of the reduce-scatter is $\alpha \cdot O(\log p) + (\beta+\gamma) \cdot O(n)$.
In the case of all-reduce and reduce-scatter, the computational cost is typically dominated by the bandwidth cost because $\beta \gg \gamma$.

@@

As shown by Kuang and Park \cite{KP13}, determining which of the four possible solutions is feasible and optimal can be done efficiently by exploiting the following properties:
\begin{itemize}
	\item if the solution to the unconstrained least squares problem admits two positive values, it is the optimal solution to the nonnegatively constrained problem,
	\item if $\M{W}$ and $\M{A}$ are both nonnegative, then the candidate solution with two zero values is never (uniquely) \LM{why uniquely?} optimal and can be discarded, and
	\item if the solution to the unconstrained problem does not admit two positive values, the better of the two remaining solutions can be determined by comparing $\V{a}_{j}^\Tra\V{w}_1 / \|\V{w}_1\|$ and $\V{a}_{j}^\Tra\V{w}_2 / \|\V{w}_2\|$.
\end{itemize}
If the unconstrained problem is solved via the normal equations, then the temporary matrices computed for the normal equations ($\M{W}^\Tra\M{W}$ and $\M{A}^\Tra\M{W}$) can be re-used to determine the better of the two solutions with a single positive variable.

@@

Given that the computational complexity of \Cref{alg:r2nnls} is $O(n)$ (or $O(m)$ when computing $\M{W}$), and the complexity of computing $\M{W}^\Tra\M{W}$ and $\M{H}^\Tra\M{H}$ is $O(m+n)$, the typical dominant cost of each iteration of Rank-2 NMF is that of computing $\M{A}^\Tra\M{W}$ and $\M{A}\M{H}$, which is $O(mn)$ when $\M{A}$ is dense and $O(\text{nnz}(A))$ when $\M{A}$ is sparse.
\LM{should our analysis keep sparse even when we don't have a sparse dataset?}

@@

\LM{should we have a little bit more of the description of the scoring method? also the language here is awkward}
The splitting procedure is specified in \Cref{alg:split}.
After the Rank-2 NMF is performed, the $\M{H}$ factor is used to determine part membership, and the columns of the $\M{W}$ factor are assigned to the child nodes.
The score of the node is computed as the reduction in overall NMF error if the node is split, which can be computed from the principal singular values of the subsets of columns of the node and its children.
The principal singular values of the children are computed via the power method (the principal singular value of the node was computed when its parent's score was computed).

@@

Given the 1D distribution, only one communication collective is required for the pair of matrix-vector multiplications.
That is, the approximate right singular vector $\V{v}$ is redundantly owned on each processor, and the approximate left singular vector $\V{u}$ is distributed across processors.
Each processor can compute its local $\V[\hat]{u}$ from $\V{v}$ without communication and use the result for its contribution to $\V{v}=\M{A}^\Tra \V{u}$.
An all-reduce collective is used to obtain a copy of $\V{v}$ on every processor for the next iteration, and the norm is redundantly computed without further communication.
Again, we use a fixed number of iterations for benchmarking, but the more practical stopping criterion is based on the relative change in $\sigma$.
\LM{we did not use a fixed number of iterations for benchmarking}

@@

To analyze the overall cost of the hierarchical clustering algorithm, we sum the costs over all nodes in the tree.
Because the shape of the tree is data dependent and affects the overall costs, for the sake of analysis we will analyze only complete levels.
The number of rows in any node is $m$, the same as the root node, as each splitting corresponds to a partition of the columns.
Furthermore, because each split is a partition, every column of $\M{A}$ is represented exactly once in every complete level of the tree.
If we assume that all nodes perform the same number of NMF iterations ($\imath$) and power method iterations $(\jmath)$, then the dominating costs of a node with $\overline{n}$ columns is
\begin{equation*}
\label{eq:nodecost}
\begin{split}
\gamma \cdot O\left( \frac{(\imath+\jmath) m\overline{n} + \imath (m+\overline{n}) }{p} \right) + \beta \cdot O((\imath+\jmath) \overline{n}) \\ + \alpha \cdot O((\imath+\jmath) \log p).
\end{split}
\end{equation*}
Because the sum of the number of columns across any level of the tree is $n$, the cost of the $\ell$th level of the tree is
\begin{equation}
\label{eq:levelcost}
\begin{split}
\gamma \cdot O\left( \frac{(\imath+\jmath) mn + \imath m 2^\ell}{p} \right) + \beta \cdot O((\imath+\jmath) n) \\ + \alpha \cdot O((\imath+\jmath) 2^\ell \log p).
\end{split}
\end{equation}
Note that the only cost that depends on the level index $\ell$ is the latency cost.
\LM{both the computation and the latency dpend on this}
Summing over levels and assuming the tree is nearly balanced and has height $O(\log k)$ where $k$ is the number of frontier nodes, we obtain an overall cost of \Cref{alg:hiernmf2} of
\begin{equation}
\label{eq:treecost}
\begin{split}
\gamma \cdot O\left( \frac{(\imath+\jmath) mn}{p}  \log k + \frac{\imath mk}{p} \right) + \beta \cdot O((\imath+\jmath) n \log k) \\ + \alpha \cdot O((\imath+\jmath) k \log p).
\end{split}
\end{equation}
